---
title: "Agricultural Land Use Analysis"
author: Ashley Alfaro
format: html
editor: visual
---

```{r}
library(ggplot2)
library(tidyverse)
library(sf)
library(here)
library(mapview)
library(RColorBrewer)
```

### Read in Data

GeoJSON files of the state of Hawaii's agricultural land use in 2015 and 2020.

```{r}
ag_2015 <- st_read(here("data/Agricultural_Land_Use_2015_Baseline.geojson"))
ag_2020 <- st_read(here("data/Agricultural_Land_Use_2020_Update.geojson"))
```

```{r}
unique(ag_2015$cropcatego) # 15 categories total
```

```{r}
cropcatego_colors <- colorRampPalette(brewer.pal(12, "Set3"))(15)
```

### Make a Map

```{r}
mapview(ag_2015, zcol = "cropcatego") # , col.regions()
```

```{r}
mapview(ag_2020, zcol = "crops_2020")
```

### Distribution of Acreage

Note: `scales = "free_y"` is used so that every graph plotted has their own unique y-axis range.

```{r}
ggplot(ag_2015, aes(x = acreage)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~island, scales = "free_y") 
  #xlim(0, 1000)
```

**Question 1:** Did we add / lose agricultural land acreage between 2015 and 2020?

```{r}
sum_2015_acreage <- sum(ag_2015$acreage) 
sum_2020_acreage <- sum(ag_2020$acreage)

sum_2020_acreage - sum_2015_acreage
```

**Question 2:** What has that land been converted to?

Analysis Ideas:

- One map of lost agricultural land between 2015 and 2020
- Agricultural land ownership
- Normalized Difference Vegitation Index (NDVI) in lost agricultural land
- Soil types
