---
title: "Isochrones"
format: html
editor: visual
---

An **isochrone** is a line or polygon on a map that connects all points reachable from a specific starting location within a set amount of time

[Inspiration from Kyle Walker](https://walker-data.com/posts/accessibility-surface/)

### Libraries

```{r}
library(mapboxapi)
library(leaflet)
library(mapview)
library(here)
library(tidyverse)
library(mapgl)
library(sf)
```

### Create Isochrone

```{r}
isos <- mb_isochrone(
  location = "One Bowerman Dr, Beaverton, OR 97005",
  profile = "driving",
  time = 1:45
)
```

```{r}
mapview(isos, zcol = "time")
```

```{r}
library(sf)
library(fasterize)

isos_proj <- st_transform(isos, 32618)

template <- raster(isos_proj, resolution = 100)

iso_surface <- fasterize(isos_proj, template, field = "time", fun = "min")

iso_surface
pal <- colorNumeric("plasma", isos$time, na.color = "transparent")

nike_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(iso_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = isos$time, pal = pal,
            title = "Drive-time from<br>Nike HQ")

nike_map
```

```{r}
nike_icon <- makeIcon(
  iconUrl = "https://nike.com/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

nike_hq <- mb_geocode("One Bowerman Dr, Beaverton, OR 97005")

nike_map %>%
  addMarkers(lng = nike_hq[1], lat = nike_hq[2], icon = nike_icon)
```

### Your turn - Recreate for Chaminade

```{r}
chaminade_isos <- mb_isochrone(
  location = "3140 Waialae Ave, Honolulu, HI 96816",
  profile = "driving",
  time = 1:20
)
```

```{r}
mapview(chaminade_isos, zcol = "time")
```

```{r}
chaminade_proj <- st_transform(chaminade_isos, 32618)

template <- raster(chaminade_proj, resolution = 100)

chaminade_surface <- fasterize(chaminade_proj, template, field = "time", fun = "min")

chaminade_surface
pal <- colorNumeric("plasma", chaminade_isos$time, na.color = "transparent")

chaminade_map <- leaflet() %>%
  addMapboxTiles(style_id = "light-v9",
                 username = "mapbox",
                 scaling_factor = "0.5x") %>%
  addRasterImage(chaminade_surface, colors = pal, opacity = 0.5) %>%
  addLegend(values = chaminade_isos$time, pal = pal,
            title = "Drive-time from<br>Chaminade University")

chamiande_icon <- makeIcon(
  iconUrl = "https://queens.org/favicon.ico",
  iconWidth = 25,
  iconHeight = 25
)

chamiande <- mb_geocode("677 ALA MOANA, Honolulu / Downtown, HI 9681")

chaminade_map |> 
  addMarkers(lng = chamiande[1], lat = chamiande[2], icon = chamiande_icon)
```

### Lets go further

Question: Can we make an accessibility surfaces map for trauma centers, ER's, hospitals, fire stations etc. Maybe combine all of them into one shiny app. Let's strategise and design...

**App Design:**

- Fire stations: mb_isochrone works well
- Ambulance:
- Shelters:
  - Can't go base off travel time for island-wide emergency situations due to possible increase in traffic
- **Flood risk zones to a flood free zone (Ashley & Faith H.)**
  - Using polygons to map high-risk flood zones and low-risk flood zones
    - [Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/HiStateGIS::tsunami-evacuation-zones/explore?location=21.410979%2C-158.148541%2C13)
    - [Extreme Tsunami Evacuation Zones](https://geoportal.hawaii.gov/datasets/extreme-tsunami-evacuation-zones/explore?location=21.410528%2C-158.148604%2C13)
  - depart_at
  - Could reference the date: July 29, 2025 \@ 2:45 PM HST — This was when statewide tsunami warning was announced
    - Should be specified as an ISO 8601 date/time, e.g. `"2025-07-29T14:45"`
  - Think about who this would be for — I think it would be for individuals who live in high risk tsunami zones
- Hospitals, trauma centers, etc.
- Police stations

## Evacuation Zones

### Read in Tsunami Zones Data

```{r}
tsunami_zones <- st_read(here("data/Tsunami_Evacuation_Zones.geojson"))
extreme_tsunami_zones <- st_read(here("data/Extreme_Tsunami_Evacuation_Zones.geojson"))
```

### Filter Polygons to Oahu

```{r}
oahu_tsunami_zones <- tsunami_zones |>
  filter(island == "OAHU")

oahu_extreme_tsunami_zones <- extreme_tsunami_zones |>
  filter(island == "OAHU")

mapview(oahu_tsunami_zones)
```

### Dissolve Each Layer Into One Polygon

There are two zones in the event of a tsunami: regular (red) and extreme (yellow). Most tsunami warnings have those in the red zones evacuate. For an extreme tsunami, those in the red zone and yellow zone need to evacuate. To make things easier, we will dissolve each layer into one polygon for the extreme scenario.

```{r}
regular_zone <- oahu_tsunami_zones |>
  st_make_valid() |>
  st_union() |>
  st_transform(4326)

extreme_zone <- oahu_extreme_tsunami_zones |>
  st_make_valid() |>
  st_union() |>
  st_transform(4326) |>
  st_union(regular_zone)
```

```{r}
mapview(extreme_zone)
```

## Address Points

- [Address Points \| City & County of Honolulu](https://honolulu-cchnl.opendata.arcgis.com/datasets/address-points-1/explore?location=21.251058%2C-158.165637%2C10&showTable=true)
- The address point layer was created to allow the city to specify which structure (or structures) on a lot has been assigned a given address (or addresses).

```{r}
oahu_address_points <- st_read(here("data/oahu_addresses_2020.geojson")) |>
  select(objectid, address_id, streetname, geocodeadd, city, zip, globalid, geometry)
```

### Flag Which Zone Each Address Falls In

We initially thought it would be best to filter it down to addresses within the tsunami zones. However, we decided to keep all addresses. That way the app can tell someone "you are not in the tsunami zone" instead of just not finding their address.

```{r}
all_addresses <- oahu_address_points |>
  st_transform(4326) |>
  mutate(
    in_regular = lengths(st_intersects(geometry, regular_zone)) > 0,
    in_extreme = lengths(st_intersects(geometry, extreme_zone)) > 0,
    label = str_squish(paste0(geocodeadd, ", ", city, ", HI ", zip)),
    lng = st_coordinates(geometry)[, 1],
    lat = st_coordinates(geometry)[, 2]
  ) |>
  st_drop_geometry() |>
  select(address_id, label, lng, lat, in_regular, in_extreme)

count(all_addresses, in_regular, in_extreme)
```

```{r}
# check for duplicate address ids
all_addresses |> count(address_id) |> filter(n > 1)
sum(duplicated(all_addresses))

# same label AND same coordinates 
sum(duplicated(all_addresses[c("label", "lng", "lat")]))
```

```{r}
# Drop exact duplicate points
all_addresses <- all_addresses |>
  distinct(address_id, .keep_all = TRUE) |>
  distinct(label, lng, lat, .keep_all = TRUE)
```

## Example Isochrone Maps

### Select an Address

```{r}
# search: returns candidate matches
find_address <- function(query, n = 20) {
  all_addresses |>
    filter(str_detect(label, fixed(query, ignore_case = TRUE))) |>
    select(address_id, label, in_regular, in_extreme) |>
    head(n)
}

# select: address_id -> c(lng, lat)
get_start <- function(id) {
  all_addresses |>
    filter(address_id == id) |>
    select(lng, lat) |>
    unlist() |>
    unname()
}
```

```{r}
find_address("waikiki")
```

### Single Time Map

```{r}
start <- get_start(2436)

iso_1445 <- mb_isochrone(
  location  = start,
  profile   = "walking",
  time      = c(5, 10, 15),
  depart_at = "2025-07-29T14:45"
) |>
  arrange(desc(time))   # largest first so smaller contours draw on top

mapview(regular_zone, color = "red", alpha.regions = 0, lwd = 2) +
  mapview(iso_1445, zcol = "time", layer.name = "Drive time (min)",
          alpha.regions = 0.4)
```

### Comparison Map (2:45 PM vs. 4:45 PM)

Swap `regular_zone` for `extreme_zone` in `zone` to view the extreme scenario.

```{r}
times <- c(5, 10, 15)
colors <- c("#d73027", "#fc8d59", "#4575b4")

get_iso <- function(start, depart_at) {
  mb_isochrone(start, profile = "driving-traffic", time = times, depart_at = depart_at) |>
    arrange(desc(time))
}

iso_1445 <- get_iso(start, "2025-07-29T14:45")
iso_1645 <- get_iso(start, "2025-07-29T16:45")

zone <- st_sf(geometry = regular_zone)   # or extreme_zone
start_pt <- st_sf(geometry = st_sfc(st_point(start), crs = 4326))

make_map <- function(iso) {
  mapboxgl(style = mapbox_style("light"), center = start, zoom = 11) |>
    add_fill_layer(id = "iso", source = iso,
                   fill_color = match_expr("time", values = times, stops = colors),
                   fill_opacity = 0.5) |>
    add_fill_layer(id = "zone", source = zone, fill_color = "black", fill_opacity = 0.3) |>
    add_circle_layer(id = "start", source = start_pt, circle_color = "black", circle_radius = 5)
}

compare(make_map(iso_1445), make_map(iso_1645), swiper_color = "green") |>
  add_legend(
    "Drive time (2:45 pm vs. 4:45 pm)",
    values = paste(times, "minutes"),
    colors = colors,
    type = "categorical",
    style = legend_style(background_opacity = 0.8)
  )
```

## Limitations of `depart_at`

I have read articles claiming it took people over an hour to leave their parking structures from their downtown offices, and how some places had three hours of traffic gridlock ([Hawaii Plans Improvements to Tsunami Response Protocol](https://www.govtech.com/em/preparedness/hawaii-plans-improvements-to-tsunami-response-protocol)). I also have secondhand anecdotes of it taking my dad an hour to get from Ewa Gentry to Waipahu. This confused me because the map shows that someone who lives near the coast in Ewa Beach can get to Waipahu in 15 minutes.

I looked into `depart_at` more and realized it does **not** reflect the actual traffic on July 29, 2025. It shows what traffic on a Tuesday at the specified time typically looks like, so it doesn't capture the gridlock that happened when the statewide tsunami warning was announced. This makes me wonder if it would be better not to use `depart_at` and have it pull traffic data from the date and time it is ran instead.
